---
title: "STAT 345 Midterm Project"
font: 12pt
date: "April 3, 2026"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

> "Our offense is like the Pythagorean Theorem. There is no answer!", *Shaquille O'Neal*

## The Background

Your role for the midterm project is that of data analyst intern at an NBA (professional basketball) team. Your direct supervisor (also part of the analytics team) has asked you to create a data visualization to illustrate how (or if) the team's shots have changed over time. After some initial clarifying questions, your supervisor confessed that they had seen some pretty cool shot charts at <http://savvastjortjoglou.com/nba-shot-sharts.html> and would like to extend the ideas a bit.

Your data for the midterm project may come from a variety of sources, including the NBA directly, as well as Basketball-Reference, HoopsHype, and others. There are several ways to access the data, but perhaps the simplest is through a package. There are a few options (each with some benefits, drawbacks, and potential headaches):

-   `nbastatR`
-   `hoopR`
-   `AdvancedBasketballStats`

```{r, echo=FALSE}
library(ggplot2)
library(hoopR)
library(gganimate)
library(dplyr)
library(gifski)
library(tidyverse)
```

```{r, echo=FALSE}

data <- load_nba_pbp(seasons = 2016:2026)
wolves_shots<- data %>% filter(team_id == "16")
```

Note: if you would prefer to analyze college basketball, packages exist for that too (such as `cbbdata`). If you would prefer to look into a different sport, we can investigate that as well. Please let me know in advance if you intend to go down one of these alternate routes.

## The Tasks

1.  (30 points) Produce a graphic displaying the shot locations for a particular team over several years. Some notes:

    -   Colors should be chosen to reflect the team, if possible.
    -   There are likely many overlaid points -- handle this by either binning these by location, or use opacity and size.
    -   Incorporate information about whether the shot was made or not (shape, color, etc.).
    -   The graphic should be well-labeled, titled, etc.
    -   Start with a graph for a single year, then extend to several years. Up to 20 years of shot data is available. Either facet these by year or animate using the years.
    -   You'll want to figure out what the coordinates mean somehow. This might be through the documentation, but could also be determined using aspects of the data itself and the dimensions of an NBA court. #coordinate_x: left/right (≈ −25 to 25) #coordinate_y: distance from hoop (0 at basket → \~47 half court) #Hoop is at (0, 0) #We flip the y-axis so the hoop appears at the top of the plot
    -   Put a basketball court on the background of the image (you'll need to scale it appropriately).

```{r, echo=FALSE}
#for getting no code to show, if errors show then do warnings= FALSE
```

```{r, echo=FALSE, warning=FALSE}
data26 <- wolves_shots %>% mutate(made = ifelse(scoring_play, "Made", "Missed"))
draw_court <- ggplot(data=data.frame(x=1,y=1),aes(x,y))+
# got the court from https://gist.github.com/edkupfer/6354404
geom_path(data=data.frame(x=c(-25,-25,25,25,-25),y=c(0,47,47,0,0)))+

geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(19+sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2))
),aes(x=x,y=y))+

geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(19-sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2)
),linetype='dashed'),aes(x=x,y=y))+

geom_path(data=data.frame(x=c(-8,-8,8,8,-8),y=c(0,19,19,0,0)))+
geom_path(data=data.frame(x=c(-6,-6,6,6,-6),y=c(0,19,19,0,0)))+

geom_path(data=data.frame(
  x=c(-4000:(-1)/1000,1:4000/1000),
  y=c(5.25+sqrt(4^2-c(-4000:(-1)/1000,1:4000/1000)^2)
)),aes(x=x,y=y))+

geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(47-sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2)
)),aes(x=x,y=y))+

geom_path(data=data.frame(
  x=c(-750:(-1)/1000,1:750/1000,750:1/1000,-1:-750/1000),
  y=c(
    c(5.25+sqrt(0.75^2-c(-750:(-1)/1000,1:750/1000)^2)),
    c(5.25-sqrt(0.75^2-c(750:1/1000,-1:-750/1000)^2))
  )
),aes(x=x,y=y))+

geom_path(data=data.frame(x=c(-3,3),y=c(4,4)),lineend='butt')+

geom_path(data=data.frame(
  x=c(-22,-22,-22000:(-1)/1000,1:22000/1000,22,22),
  y=c(
    0,169/12,
    5.25+sqrt(23.75^2-c(-22000:(-1)/1000,1:22000/1000)^2),
    169/12,0
  )
),aes(x=x,y=y))+

coord_fixed(xlim = c(-25, 25), ylim = c(0, 47)) +

geom_jitter(
  data = data26,
  aes(
    x = coordinate_y,
    y = 47 - coordinate_x,
    color = made,
    shape = made
  ),
  width = 0.5,
  height = 0.5,
  alpha = 0.3,
  size = 1.5
) +

scale_color_manual(
  values = c("Made" = "#236192", "Missed" = "#78BE20")) +
transition_states(season, transition_length = 5, state_length = 75)+
#creates 
labs(
  title = "Minnesota Timberwolves Shot Chart {closest_state}",
  subtitle = "Shot locations with Makes vs. Misses",
  color = "Shot Result",
  shape = "Shot Result"
)

draw_court
```

2.  (30 points) Summarize the graphic/series of graphics into a digestible, bullet-point brief report for front-office staff. Some notes:

    -   The main body of the report should be very brief -- just the graphic(s) and the bullet-pointed list of findings, which should be short and clear.
    -   Include a more detailed explanation of these bullet points, for further reading by those interested. This section should follow the bullet-point section, but should be organized similarly for reference.
    -   Your report to the front-office shouldn't include any code.
    -   This report should be generated using RMarkdown. However, the choice of output type (Word, PDF, or HTML) is up to you (you could even make slides if you want to).

## **Key Findings:**

-   **High Output Areas:** Most made shots are concentrated around the corners of the 3-point line, near the basket, and at the free throw line.

-   **Low Output Areas:** Mid range shots show that they aren't made very frequently.

-   **Seasonal Trends:** Shooting efficiency and volume has increased over the 2016-2026 seasons.

-   **3-Point Reliance:** There is an increase in the rate of 3 point attempts made in the later seasons.

-   **Shot Clustering:** Some players may favor from shooting from a certain spot, they also might have similar plays that cause these clusters.

## **Detailed Explanation:**

-   **High Output Areas:** The chart indicates that the Timberwolves consistently score a higher percentage of shots near the basket and at corner three positions.These zones are strategic for designing offensive sets and prioritizing high-value attempts. Focusing plays and practice drills around these hotspots could maximize scoring efficiency and optimize game-time performance.

-   **Low Output Areas:** Mid-range jumpers show lower success rates, suggesting that resources and practice may be better focused elsewhere. Encouraging shot selection closer to the rim or beyond the 3-point arc could improve overall scoring efficiency. Limiting mid-range attempts in game plans may also reduce wasted possessions and improve team shooting percentages.

-   **Seasonal Trends:** Comparing multiple seasons shows moderate improvement in shooting accuracy in some areas, with a gradual increase in three-point attempts over time. These trends can help identify player development patterns and inform off-season training plans.Looking at these patterns supports data-driven decisions on player rotations and can help to target skill development throughout the season.

-   **3-Point Reliance:** Later seasons indicate an increased proportion of three-point shots, reflecting broader league trends and team strategy. This has implications for spacing, lineups, and defensive match-ups. Spending more time practicing 3-point shoots could further exploit these trends, which could give the team a strategic advantage over opponents.

-   **Shot Clustering:** Specific court positions are repeatedly targeted by players, highlighting tendencies that could be leveraged or defended against. Identifying these clusters can help with opponent scouting and game planning. Using this information can help us plan both offense and defense around high-traffic shooting areas.

3.  (30 points) Write and document clean, efficient, reproducible code. Some notes:

    -   This code will be viewed by your direct supervisor.
    -   The code file should include your code to gather, join, and clean the data; the code to generate the graphic(s) presented; and your commentary on the results (so, a single .rmd file, or an .rmd file that sources an .r file).
    -   Your code should be clean, organized, and reproducible. Remove unnecessary/scratch/exploratory code.
    -   Your code should be well commented. In particular, any decisions or judgement calls made in the analysis process should be explained/justified. Sections of code should be identified even if not functionalized (including purpose, data/argument inputs, analysis outputs).

```{r, echo=TRUE, warning=FALSE}
#load in libraries 
library(ggplot2)
library(hoopR)
library(gganimate)
library(dplyr)
library(gifski)
library(tidyverse)
#load in the data for all seasons we are looking for
data <- load_nba_pbp(seasons = 2016:2026)
#Filter for the Minnesota Timberwolves (team_id = 16)
wolves_shots<- data %>% filter(team_id == "16")
#Creates a 'made' column to label made vs missed shots
data26 <- wolves_shots %>% mutate(made = ifelse(scoring_play, "Made", "Missed"))
# got the court from https://gist.github.com/edkupfer/6354404
draw_court <- ggplot(data=data.frame(x=1,y=1),aes(x,y))+
  #Court boundary rectangle
geom_path(data=data.frame(x=c(-25,-25,25,25,-25),y=c(0,47,47,0,0)))+
#top of free point arcs
geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(19+sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2))
),aes(x=x,y=y))+
#free throw arcs
geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(19-sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2)
),linetype='dashed'),aes(x=x,y=y))+
#paint rectangle (outer and inner)
geom_path(data=data.frame(x=c(-8,-8,8,8,-8),y=c(0,19,19,0,0)))+
geom_path(data=data.frame(x=c(-6,-6,6,6,-6),y=c(0,19,19,0,0)))+
#Free throw circle (top arc)
geom_path(data=data.frame(
  x=c(-4000:(-1)/1000,1:4000/1000),
  y=c(5.25+sqrt(4^2-c(-4000:(-1)/1000,1:4000/1000)^2)
)),aes(x=x,y=y))+
#Three-point line arc (baseline to arc)
geom_path(data=data.frame(
  x=c(-6000:(-1)/1000,1:6000/1000),
  y=c(47-sqrt(6^2-c(-6000:(-1)/1000,1:6000/1000)^2)
)),aes(x=x,y=y))+
#Smaller inner circle around basket (rim area)
geom_path(data=data.frame(
  x=c(-750:(-1)/1000,1:750/1000,750:1/1000,-1:-750/1000),
  y=c(
    c(5.25+sqrt(0.75^2-c(-750:(-1)/1000,1:750/1000)^2)),
    c(5.25-sqrt(0.75^2-c(750:1/1000,-1:-750/1000)^2))
  )
),aes(x=x,y=y))+
#Backboard line
geom_path(data=data.frame(x=c(-3,3),y=c(4,4)),lineend='butt')+
#Restricted area/three-point extended line
geom_path(data=data.frame(
  x=c(-22,-22,-22000:(-1)/1000,1:22000/1000,22,22),
  y=c(
    0,169/12,
    5.25+sqrt(23.75^2-c(-22000:(-1)/1000,1:22000/1000)^2),
    169/12,0
  )
),aes(x=x,y=y))+
#Set fixed ratio to prevent distortion
coord_fixed(xlim = c(-25, 25), ylim = c(0, 47)) +
#Plotting the shot attempts using jitter for better visiblity, color and shapre represent made/missed shots. Width, alpha, height and size controls the appearence of the points
geom_jitter(
  data = data26,
  aes(
    x = coordinate_y,
    y = 47 - coordinate_x,
    color = made,
    shape = made
  ),
  width = 0.5,
  height = 0.5,
  alpha = 0.3,
  size = 1.5
) +
#Maunal color scale to match made and missed labels  as well as team colors
scale_color_manual(
  values = c("Made" = "#236192", "Missed" = "#78BE20")) +
  #Animates the charts by season for visualization over the years
transition_states(season, transition_length = 5, state_length = 75)+
#Chart labels
labs(
  title = "Minnesota Timberwolves Shot Chart {closest_state}",
  subtitle = "Shot locations with Makes vs. Misses",
  color = "Shot Result",
  shape = "Shot Result"
)
#Shows the chart just made
draw_court
```

4.  (10 points) Above and Beyond. Choose either option below. You are welcome to explore both, but only one is required.

-   Option 1: Explore the data a bit, and create a graphic that uses (or incorporates) different information than what was used above. Some notes:
    -   Create an additional graphic that incorporates at least one additional variable not previously used (this should add to the graphic in part 1). The additional data should be drawn from a different dataset (function call) than the original graphic used. These two (or more) datasets may need to be joined appropriately.
    -   You can either add more information to the plot above, or create a different plot.
    -   Formatting, labelling, etc. are all important here too.
    -   Adding marginal densities or other "bells and whistles" might offer additional insight.
    -   This graphic should be included at the end of the report (after the more detailed explanations).
    -   You should include a brief description of the graphic (highlighting the different/additional information used).
-   Option 2: If the NBA were to incorporate a 4-point shot, where would you draw a 4-point arc? Some notes:
    -   You likely should base your decision at least partly on proportion of shots made from a given range. You might consider an expected value calculation here.
    -   Your arc can be shaped as you see fit; simple arcs are sufficient for this exploration.
    -   Provide an example of a consequence (positive or negative) if a 4-point shot was incorporated. (e.g., "my_favorite_player's season point total would increase by x%")
    -   You do not need to create a plot representing your arc, though you are welcome to do so!

```{r}
distance_summary <- data26 %>% mutate(distance= sqrt((coordinate_x)^2 + (coordinate_y)^2)) %>% summarise(median_distance= median(distance), line_3pt= mean(distance >= 23.75)) # 3-pt line is approx. 23.75 ft
distance_summary
```

This tells us that the median shot distance for the Timberwolves is 38.75 feet. With 88.02% of the shots already being inside the 3-point range, which means 11.98% of shots are being taken beyond the 3-point line. If we were to create a 4-point line, I would suggest it be around the 28-30 feet range due to our 38.75 median distance which could be from so buzzer-beating shots attempted from past half-court.

With my favorite player being Jaden McDaniels, if he currently attempts 50 shots per season from beyond my proposed 4-point line (between 28-30 feet). If his shooting accuracy from this distance is 40%, then his current expected points (if its a 3-point shot) will be 60 points, but his expected points if its a 4-point shot would be 80 points. Adding the 4-point line would positvely impact McDaniels by 33%, which encourages more attempts from farther away and potentially alters offensive strategies. This could also be a negative impact because he might try and take more long shots that he would have a low chance of making in hopes for an extra point. This could overall reduce efficiency if players were extending beyond their usual range.

## The Deliverables

1.  Upload your report and code file(s) to GitHub by 5:00pm on Friday, April 3.
2.  Submit (on Canvas) your report, code, and link to your GitHub repository by 5:00pm on Friday, April 3
